Childhood exposure to non-persistent endocrine disruptors, glucocorticosteroids, and attentional function: A study based on the parametric g-formula

### 2024-03-30 output/paper/manuscript.qmd
Comments from MV v2.

The impact of exposure to {{< acr edcs >}} on human health has attracted considerable research interest. While research in this area has mainly investigated the effects of prenatal exposure on child neurodevelopment [@RamirezGalvez-OntiverosGonzalez-Domenech:2022], little is still known about childhood exposure. In this study, consisting of 1,297 children from 6 European birth cohorts, we observed that short-term childhood exposure to certain non-persistent {{< acr edcs >}} was associated with attentional function (MEPA, MEHP, oh-MiNP, and oxo-MiNP), and with total production of cortisol, cortisone, and corticosterone (DEP, DMP, DMTP, BPA, ETPA, MEPA, MEHP, oh-MiNP, and oxo-MiNP). Increased production of these glucocorticosteroids did not seem to affect attentional function. Some of these effects differed for females and males, including significant differences for the effects of increased production of cortisone and corticosterone on {{< acr hitrtse >}}. Specifically, an increased production of these glucocorticosteroids was associated with lower values of {{< acr hitrtse >}} for females, and higher values for males. Taken together, these results suggest that these non-persistent {{< acr edcs >}} might be responsible for perturbations of the {{< acr hpa >}} axis' homeostasis, and that higher levels of these glucocorticosteroids might interfere with different functions of attention in a sex-specific manner.

To the best of our knowledge, no other study has investigated the effects of childhood exposure to multiple classes of non-persistent {{< acr edcs >}} in relation to attentional function.
More generally, the literature on non-persistent {{< acr edcs >}} and neurodevelopment in children has mostly focused on {{< acr opps >}} [@BouchardBellingerWright:2010; @Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @CartierWarembourgLeManer-Idrissi:2016; @YuDuChiou:2016], phthalate metabolites [@HuangChenSu:2015; @HuangTsaiChen:2017; @KimHongShin:2017; @LiPapandonatosCalafat:2019; @BalalianWhyattLiu:2019; @ShoaffCoullWeuve:2020; @JankowskaPolanskaHanke:2019; @VilmandBeckBilenberg:2023], and {{< acr bpa >}} [@TewarAuingerBraun:2016; @LiZhangKuang:2018; @Rodriguez-CarrilloMustielesPerez-Lobato:2019].
González-Alzaga et al. and Cartier et al. evaluated cross-sectional associations between {{< acr dap >}} metabolites and subtests of the Wechsler Intelligence Scale for Children [@WechslerKodama:2003] in European children with ages between 6 and 11 years. Higher levels of {{< acr dap >}} metabolites (DMP, DMTP, DMDTP, DEP, DETP, and DEDTP) were associated with lower scores of {{< acr iq >}} and verbal comprehension, especially in boys [@Gonzalez-AlzagaHernandezRodriguez-Barranco:2015], while higher levels of diethylphosphate metabolites (DEP, DETP, DEDTP) were associated with lower working memory scores [@CartierWarembourgLeManer-Idrissi:2016].
There is also preliminary evidence of a possible association between exposure to certain {{< acr opps >}} and {{< acr adhd >}} in children [@BouchardBellingerWright:2010; @YuDuChiou:2016]. Specifically, Bouchard et al. found evidence of a cross-sectional association between dimethyl alkylphosphate metabolites (DMP, DMTP, and DMDTP) and {{< acr adhd >}} in children aged 8 to 15 years from {{< acr nhanes >}}, while Yu et al. found a dose-response relationship between DMP and {{< acr adhd >}} in Taiwanese children aged 4 to 15 years.
Preliminary evidence is also available for several phthalate metabolites in relation to cognitive development in childhood. Higher levels of di(2-ethylhexyl) phthalate metabolites (including MEHP, MEHHP, and MEOHP) were associated with lower intelligence scores in children aged 2 to 12 years [@HuangChenSu:2015], lower scores of {{< acr iq >}} and verbal intelligence, more omission errors (a measure of inattention), and higher scores of response time variability (a measure of sustained attention) in 6-year old Korean children [@KimHongShin:2017], poorer fine motor skills in preadolescent boys [@BalalianWhyattLiu:2019], and lower intelligence scores in 7-year old children [@VilmandBeckBilenberg:2023]. Further associations were found for MEOHP with lower scores of {{< acr iq >}} [@HuangChenSu:2015] and verbal intelligence in Taiwanese children aged 6 to 12 years [@HuangTsaiChen:2017], and for dibutyl phthalate metabolites (MnBP and MiBP) with impaired verbal intelligence [@HuangTsaiChen:2017]. There is further preliminary evidence that associations between certain phthalate metabolites and cognitive abilities vary by timing of exposure assessment [@LiPapandonatosCalafat:2019].
Among phenols, some studies provide preliminary evidence of an association between BPA and {{< acr adhd >}} in children aged 8 to 15 years [@TewarAuingerBraun:2016] and in a case-control study of children aged 6 to 12 years [@LiZhangKuang:2018], especially in boys. Except for working memory, there does not seem to be evidence of an association between BPA and cognitive abilities in Spanish boys aged 9 to 11 years [@Rodriguez-CarrilloMustielesPerez-Lobato:2019].
Few studies have looked into different classes of non-persistent {{< acr edcs >}}. Shoaff et al., for instance, investigated cross-sectional associations between multiple {{< acr edcs >}} and {{< acr adhd >}}-related behaviors in 15-year old adolescents, finding a higher risk of {{< acr adhd >}}-related behavior problems with higher levels of antiandrogenic phthalate metabolites (MEHP, MEHHP, MEOHP, MECPP, MnBP, MiBP, MBzP, monohydroxyisobutyl phthalate (MHiBP), monocarboxyoctyl phthalate (MCOP), monoisononyl phthalate (MNP), and monohydroxybutyl phthalate (MHBP)), especially in boys [@ShoaffCoullWeuve:2020].

We are not aware of other epidemiological studies investigating childhood exposure to phthalates metabolites, phenols, and {{< acr opps >}}, in relation to urinary glucocorticosteroid levels in childhood. Prior epidemiological research provides preliminary evidence for an association between certain non-persistent {{< acr edcs >}} with higher levels of glucocorticoids [@KimLeeMoon:2018; @SunLiJin:2018; @SearsLiuLanphear:2023]. Repeated measures up to 15 months of age of the phthalate metabolites MEHHP, MEOHP, MiBP, and MnBP showed positive associations with free cortisol in Korean children, with no effect modification by sex [@KimLeeMoon:2018]. In a cohort of Chinese pregnant women, phthalate metabolites were measured at 14, 24, and 36 weeks of gestation, and the glucocorticoids cortisol and cortisone were measured in cord blood. Third-trimester levels of MEHP were positively associated with cortisol, while MECPP and MEOHP were negatively associated with cortisone [@SunLiJin:2018]. Time- and chemical-dependent sex differences were also found: during the third trimester, MEHHP and MEOHP were positively associated with cortisol in females, while negatively associated in males [@SunLiJin:2018]. In a longitudinal study, a mixture of several phthalate metabolites, driven by MEP, MiBP, and MBzP, measured in childhood, showed a positive association with hair cortisol measured at 12 years of age [@SearsLiuLanphear:2023]. While in the present study we did find positive {{< acr mcs >}} between some phthalate metabolites (MEHP, oh-MiNP, and oxo-MiNP) and the glucocorticosteroids, there are important differences with the previous studies. First, exposure assessment was performed during gestation [@SunLiJin:2018] or the first 15 months of life [@KimLeeMoon:2018], not during childhood. Second, the glucocorticosteroids were measured in other matrices, specifically in cord blood [@SunLiJin:2018] or hair [@SearsLiuLanphear:2023]. Finally, @SearsLiuLanphear:2023 investigated mixture effects. Contrary to these studies [@KimLeeMoon:2018; @SearsLiuLanphear:2023], we did find effect modification by sex.

>>>>

In this study, consisting of 1,297 children from 6 European birth cohorts, we observed that short-term childhood exposure to certain non-persistent {{< acr edcs >}} was associated with attentional function (MEPA, MEHP, oh-MiNP, and oxo-MiNP), and with total production of cortisol, cortisone, and corticosterone (DEP, DMP, DMTP, BPA, ETPA, MEPA, MEHP, oh-MiNP, and oxo-MiNP). Increased production of these glucocorticosteroids did not seem to affect attentional function. Some of these associations differed for females and males, including significant differences in associations of increased production of cortisone and corticosterone with {{< acr hitrtse >}}. Taken together, these results suggest that exposure to non-persistent {{< acr edcs >}} might be responsible for perturbations of the {{< acr hpa >}} axis' homeostasis, and that higher levels of glucocorticosteroids might interfere with different functions of attention in a sex-specific manner.

To the best of our knowledge, no other study has investigated the effects of childhood exposure to multiple classes of non-persistent {{< acr edcs >}} in relation to attentional function.
More generally, the literature on childhood exposure to non-persistent {{< acr edcs >}} and other neurodevelopment outcomes in children has mostly focused on {{< acr opps >}} [@BouchardBellingerWright:2010; @Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @CartierWarembourgLeManer-Idrissi:2016; @YuDuChiou:2016], phthalate metabolites [@HuangChenSu:2015; @HuangTsaiChen:2017; @KimHongShin:2017; @LiPapandonatosCalafat:2019; @BalalianWhyattLiu:2019; @ShoaffCoullWeuve:2020; @JankowskaPolanskaHanke:2019; @VilmandBeckBilenberg:2023], and {{< acr bpa >}} [@TewarAuingerBraun:2016; @LiZhangKuang:2018; @Rodriguez-CarrilloMustielesPerez-Lobato:2019].
González-Alzaga et al. and Cartier et al. evaluated cross-sectional associations between {{< acr dap >}} metabolites and subtests of the Wechsler Intelligence Scale for Children [@WechslerKodama:2003] in European children with ages between 6 and 11 years. Higher levels of {{< acr dap >}} metabolites (DMP, DMTP, DMDTP, DEP, DETP, and DEDTP) were associated with lower scores of {{< acr iq >}} and verbal comprehension, especially in boys [@Gonzalez-AlzagaHernandezRodriguez-Barranco:2015], while higher levels of diethylphosphate metabolites (DEP, DETP, DEDTP) were associated with lower working memory scores [@CartierWarembourgLeManer-Idrissi:2016].
There is also preliminary evidence of a possible association between exposure to certain {{< acr opps >}} and {{< acr adhd >}} in children [@BouchardBellingerWright:2010; @YuDuChiou:2016]. Specifically, Bouchard et al. found evidence of a cross-sectional association between dimethyl alkylphosphate metabolites (DMP, DMTP, and DMDTP) and {{< acr adhd >}} in children aged 8 to 15 years from {{< acr nhanes >}}, while Yu et al. found a dose-response relationship between DMP and {{< acr adhd >}} in Taiwanese children aged 4 to 15 years.

Preliminary evidence is also available for several phthalate metabolites in relation to cognitive development in childhood. Higher levels of di(2-ethylhexyl) phthalate metabolites (including MEHP, MEHHP, and MEOHP) were associated with lower intelligence scores in children aged 2 to 12 years [@HuangChenSu:2015], lower scores of {{< acr iq >}} and verbal intelligence, more omission errors (a measure of inattention), and higher scores of response time variability (a measure of sustained attention) in 6-year old Korean children [@KimHongShin:2017], poorer fine motor skills in preadolescent boys [@BalalianWhyattLiu:2019], and lower intelligence scores in 7-year old children [@VilmandBeckBilenberg:2023]. Further associations were found for MEOHP with lower scores of {{< acr iq >}} [@HuangChenSu:2015] and verbal intelligence in Taiwanese children aged 6 to 12 years [@HuangTsaiChen:2017], and for dibutyl phthalate metabolites (MnBP and MiBP) with impaired verbal intelligence [@HuangTsaiChen:2017]. There is further preliminary evidence that associations between certain phthalate metabolites and cognitive abilities vary by timing of exposure assessment [@LiPapandonatosCalafat:2019].
Finally, few studies have looked into different classes of non-persistent {{< acr edcs >}}. Shoaff et al., for instance, investigated cross-sectional associations between multiple {{< acr edcs >}} and {{< acr adhd >}}-related behaviors in 15-year old adolescents, finding a higher risk of {{< acr adhd >}}-related behavior problems with higher levels of antiandrogenic phthalate metabolites, especially in boys [@ShoaffCoullWeuve:2020]. Our findings, indicating that short-term childhood exposure to certain phthalate metabolites (MEHP, oh-MiNP, and oxo-MiNP) was associated with attentional function, adds to this growing evidence base suggesting that childhood phthalate exposure may impact child neurodevelopment.

Among phenols, some studies provide preliminary evidence of an association between BPA and {{< acr adhd >}} in children aged 8 to 15 years [@TewarAuingerBraun:2016] and in a case-control study of children aged 6 to 12 years [@LiZhangKuang:2018], especially in boys. Except for working memory, there was no evidence of an association between BPA and cognitive abilities in Spanish boys aged 9 to 11 years [@Rodriguez-CarrilloMustielesPerez-Lobato:2019]. We did not observe an association between BPA and attention function in the present study, but this study is the first to suggest that childhood paraben (MEPA) exposure may be associated with attention function.

We are not aware of other epidemiological studies investigating childhood exposure to phthalates metabolites, phenols, and {{< acr opps >}}, in relation to urinary glucocorticosteroid levels in childhood. However, prior epidemiological research provides preliminary evidence for an association between certain non-persistent {{< acr edcs >}} measured at other time points with higher levels of glucocorticoids measured in other biological matrices [@KimLeeMoon:2018; @SunLiJin:2018; @SearsLiuLanphear:2023]. Repeated measures up to 15 months of age of the phthalate metabolites MEHHP, MEOHP, MiBP, and MnBP showed positive associations with free cortisol in urine in Korean children, with no effect modification by sex [@KimLeeMoon:2018]. In a cohort of Chinese pregnant women, phthalate metabolites were measured at 14, 24, and 36 weeks of gestation, and the glucocorticoids cortisol and cortisone were measured in cord blood. Third-trimester levels of MEHP were positively associated with cortisol, while MECPP and MEOHP were negatively associated with cortisone [@SunLiJin:2018]. Time- and chemical-dependent sex differences were also found: during the third trimester, MEHHP and MEOHP were positively associated with cortisol in females, while negatively associated in males [@SunLiJin:2018]. In a longitudinal study, a mixture of several phthalate metabolites, driven by MEP, MiBP, and MBzP, measured in childhood, showed a positive association with hair cortisol measured at 12 years of age [@SearsLiuLanphear:2023]. Our findings also indicate associations between certain phthalate metabolites (MEHP, oh-MiNP, and oxo-MiNP) and glucocorticosteroids, but differences in the exposure assessment time points and in the biological matrices used for glucocorticosteroids determinations make a direct comparison difficult. Further, we did find effect modification by sex.
